### 2018-10-14 Manuscript/Manuscript.tex
Additions to discussion and methods

Yuem Park\textsuperscript{1},
Nicholas L. Swanson-Hysell\textsuperscript{1},
Francis A. Macdonald\textsuperscript{2}
Lorraine E. Lisiecki\textsuperscript{2}

@@

Outlines of large igneous provinces (LIPs) through the Phanerozoic (Fig. \ref{fig:LIP_map}) were taken from the compilation of \cite{Ernst2017a}, and emplacement ages were taken from the literature (Fig. \ref{fig:LIP_map}; Table \ref{tab:LIPs}). The compilation of \cite{Ernst2017a} seeks to reconstruct the original areal extent of LIPs with the caveat that there can be significant uncertainty with doing so, particularly for older more deeply eroded LIPs. These polygons encapsulate all of the rocks associated with a given LIP including dikes and sills (Fig. \ref{fig:LIP_map}). For some LIPs, this approach may lead to an overestimate of original areal extent given that subsurface intrusions could extend over a broader area than the surface volcanics. However, such intrusions likely provide the best estimate available of original extent for ancient LIPs. The extent of presently exposed volcanics associated with the LIPs were taken from a number of resources including the PLATES compilation (Table \ref{tab:LIPs}; \citealp{Coffin2006a}) and more recent compilation efforts associated with the Large Igneous Provinces Commission.

After LIPs are emplaced they will progressively erode. In order to account for this decrease in area through with time, \cite{Godderis2017a} took the approach of fitting estimates of the original surface extent and estimates of the current extent for five LIPs with an exponential decay function. They used the resulting exponential decay constants to develop a first-order model to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which we have estimates of the original areal extent of the province and the current areal extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate approximate model for the progressive reduction in LIP area. The best-fit exponential function results in a half-life of LIP area of 36 Myr. There is variability in the individual half-lives for LIPs such that they range from shorter half-lives of $\sim$20 Myr up to $\sim$120 Myr for the Kalkarindji and Deccan Traps. In the analysis of LIP area through time, we implement decay scenarios informed by this analysis. `Decay 1 scenario' uses the best-fit half-life of 38 Myr while `decay 2 scenario' uses the slower decay with a half-life of $\sim$120 Myr. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis. Of the tectonic factors that could alter exposed LIP area, the most consequential is near-immediate burial of LIP volcanics by sediments if they are co-located with a rift basin. There are numerous examples in the record where there is partial or complete burial of a LIP associated with rifting and thermal subsidence (Table \ref{tab:LIPs}). For example, the Afar LIP is both associated with the Ethiopian Traps which form plateau flood basalts and successful rifting in the region of the Red Sea that has resulted in burial. To account for the rapid decrease in exposed surface area that would result from burial by sediments in a rift basin, we impose two different burial scenarios for LIPs that are co-located with rifting. `Burial 1 scenario' imposes instantaneous burial of 50$\%$ of the LIP area while `burial 2 scenario' imposes burial of the entire LIP as an end-member type scenario. The LIP area analysis uses all of the available combinations of these distinct decay and burial scenarios. Our LIP reconstruction differs from that of \cite{Johansson2018a} in several respects. In contrast to the decay and decay+burial scenarios implemented on estimates of original LIP extent, \cite{Johansson2018a} uses a static extent for each LIP throughout the reconstruction with some of the polygons corresponding to the present-day surface extent and some of the representing the original extent that includes currently buried portions of the LIP (e.g. the North American Midcontinent Rift). The LIP reconstruction used in this study includes pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs do not significantly impact the calculated LIP areas throughout the Phanerozoic.

The original extent LIP polygons were assigned a plate ID corresponding to a tectonic unit on Earth using the polygons of \cite{Torsvik2016a} for the Phanerozoic. The LIP polygons were reconstructed from 520 Ma to the present along with the tectonic units utilizing the paleogeographic model of \cite{Torsvik2016a} in the spin axis reference frame (anchor plate ID of 1). This paleogeographic model was updated to include a revision to Ordovician Laurentia \citep{Swanson-Hysell2017a} and the Paleozoic of Asia \citep{Domeier2018a}. Reconstructions and area calculations within latitude bands utilized the pyGPlates function library and custom Python scripts documented within a Jupyter notebook that reproduces the analysis and development of the associated visualizations. At 5 Myr timesteps, the total LIP area and the LIP area reconstructed within the tropical rain belt (15\textdegree S to 15\textdegree N) were calculated for the various decay and burial scenarios (Fig. \ref{fig:LIP_area}).

To determine whether there is a relationship between Earth's climate state and tropical LIP area, we compared these areas to a compilation of the latitudinal extent of continental ice sheets through time over the Phanerozoic Eon \fig{}. The goal in doing so is to evaluate the hypothesis that there is a correlation between LIP area in the tropics and Earth's long-term climate state. The latitudinal extent of land ice is reflective of Earth's overall climate and allows for the delineation of glacial and non-glacial climate states. Another approach that has been implemented in the literature is to compare LIP area records to proxy compilations of pCO$_2$. Such proxy compilations are problematic as they are poorly calibrated in deep time, prone to alteration and reliant on myriad assumptions. While the latitudinal extent of land ice does provide a physical record of Earth's climate, is an imperfect tracker as it is insensitive to changes in temperatures during non-glacial intervals and is influenced by additional factors such as the physical geography of the continents during glacial intervals. To seek to deal with these issues, we take two approaches for comparison. The first to to look at the correlation between LIP area and the extent of ice away from the pole (Fig. \ref{fig:LIP_correlation}). The second is to consider the overall between intervals of high LIP area in the tropics (defined as $>30\%$ of the maximum) and intervals of glacial climate (defined as ice extent $>30\%$ of the maximum) which places less emphasis on the detailed magnitudes of the compiled ice extent and LIP area records.

@@

In the original models that proposed the ``Fire and Ice'' hypothesis as an explanation for the onset of Sturtian glaciation, chemical weathering was modeled as an Arhennius relationship where it was solely a function of temperature and run-off \citep{Donnadieu2004a}. However, such an approach neglects the effects that can result through soil shielding and regolith development in low-relief regions. Recent progress on understanding the relationships between landscapes, topography and chemical weathering (e.g. \cite{Maher2014a}) reveals that these effects are quite important. As a result, more recent modeling of chemical weathering that has incorporated such processes highlights the importance of high-relief regions relative to low-relief ones \citep{Godderis2017b}). Without active uplift, soil shielding from regolith development on low-relief LIPs could significantly decrease the local weatherability of a LIP and mute its impact on global weatherability.

DISCUSS FRANKLIN AND UMKONDO AREAS AND COMPARE TO THE PHANEROZOIC RECORD.

@@

\begin{figure}[h!]
\begin{center}
	\includegraphics[width=\textwidth]{Manuscript/Figures/LIP_Map_0_520.pdf}
	\caption{\textbf{Map of current extent of lithologies associated with LIPs that erupted between 520 Ma and the present (shown in orange) as well as the estimates of the initial LIP area used in the area analysis (shown in blue).}}
	\label{fig:LIP_map}
\end{center}
\end{figure}

>>>>

Yuem Park\textsuperscript{1},
Nicholas L. Swanson-Hysell\textsuperscript{1},
Francis A. Macdonald\textsuperscript{2},
Lorraine E. Lisiecki\textsuperscript{2}

@@

Outlines of LIPs through the Phanerozoic (Fig. \ref{fig:LIP_map}) were taken from the compilation of \citet{Ernst2017a}, and emplacement ages were taken from the literature (Fig. \ref{fig:LIP_map}; Table \ref{tab:LIPs}). The compilation of \citet{Ernst2017a} seeks to reconstruct the original areal extent of LIPs with the caveat that there can be significant uncertainty with doing so, particularly for older more deeply eroded LIPs. These polygons encapsulate all of the rocks associated with a given LIP including dikes and sills (Fig. \ref{fig:LIP_map}). For some LIPs, this approach may lead to an overestimate of original areal extent given that subsurface intrusions could extend over a broader area than the surface volcanics. However, such intrusions likely provide the best estimate available of original extent for ancient LIPs. The extent of presently exposed volcanics associated with the LIPs were taken from a number of resources including the PLATES compilation (Table \ref{tab:LIPs}; \citealp{Coffin2006a}) and more recent compilation efforts associated with the Large Igneous Provinces Commission.

After LIPs are emplaced they will progressively erode. In order to account for this decrease in area through with time, \citet{Godderis2017a} took the approach of fitting estimates of the original surface extent and estimates of the current extent for 5 LIPs with an exponential decay function. They used the resulting exponential decay constants to develop a first-order model to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which we have estimates of the original areal extent of the province and the current areal extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate approximate model for the progressive reduction in LIP area. The best-fit exponential function results in a LIP area half-life of 29~Myr. However, since we explicitly account for LIP burial separately in our models (see below), we prefer to exclude the 6 of 19 LIPs which have been partially or completely buried in our estimation of a representative LIP area half-life, which yields a slightly higher best-fit half-life of 36~Myr (Fig. \ref{fig:LIP_preservation}). There is variability in the individual half-lives for LIPs such that they range from shorter half-lives of $\sim$20~Myr up to $\sim$120~Myr for the Kalkarindji and Deccan Traps. In the analysis of LIP area through time, we implement decay scenarios informed by this analysis. Scenario `decay 1' uses the best-fit half-life of 36 Myr, while `decay 2' uses the slower decay with a half-life of $\sim$120~Myr. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis.

Of the tectonic factors that could alter exposed LIP area, the most consequential is near-immediate burial of LIP volcanics that are co-located with a rift basin by sediments. There are numerous examples in the record where there is partial or complete burial of a LIP associated with rifting and thermal subsidence (Table \ref{tab:LIPs}). For example, the Afar LIP is both associated with the Ethiopian Traps which form plateau flood basalts, and successful rifting in the region of the Red Sea that has resulted in burial. To account for the rapid decrease in exposed surface area that would result from burial by sediments in a rift basin, we impose two different burial scenarios for LIPs that are co-located with rifting. `Burial 1' imposes instantaneous burial of 50$\%$ of the LIP area while `burial 2' imposes instantaneous burial of the entire LIP as an end-member scenario. The LIP area analysis uses all of the available combinations of these distinct decay and burial scenarios. 
Our LIP reconstruction differs from that of \citet{Johansson2018a} in several respects. In contrast to the decay and decay+burial scenarios implemented on estimates of original LIP extent in this study, \citet{Johansson2018a} uses a static extent for each LIP throughout the reconstruction with some of the polygons corresponding to the present-day surface extent and some representing the original extent that includes currently buried portions of the LIP (e.g. the North American Mid-continent Rift). Furthermore, the LIP reconstruction used in this study includes pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs do not significantly impact the calculated LIP areas throughout the Phanerozoic.

The original extent LIP polygons were assigned a plate ID corresponding to a tectonic unit on Earth using the polygons of \citet{Torsvik2016a} for the Phanerozoic. The LIP polygons and tectonic units were reconstructed from 520~Ma to the present utilizing the paleogeographic model of \citet{Torsvik2016a} in the spin axis reference frame (anchor plate ID of 1). This paleogeographic model was updated to include a revision to Ordovician Laurentia \citep{Swanson-Hysell2017a} and the Paleozoic of Asia \citep{Domeier2018a}. Reconstructions and area calculations within latitude bands utilized the pyGPlates function library and custom Python scripts documented within a Jupyter notebook that reproduces the analysis and development of the associated visualizations. The total LIP area and the LIP area reconstructed within the tropical rain belt (15\textdegree~S to 15\textdegree~N) were calculated for the various decay and burial scenarios at a resolution of 5~Myr (Fig. \ref{fig:LIP_area}).

To determine whether there is a relationship between Earth's climate state and tropical LIP area, we compared these areas to a compilation of the latitudinal extent of continental ice sheets through time over the Phanerozoic Eon (Fig. \ref{fig:LIP_area}). The goal in doing so is to evaluate the hypothesis that there is a correlation between LIP area in the tropics and Earth's long-term climate state. The latitudinal extent of land ice is reflective of Earth's overall climate and allows for the delineation of glacial and non-glacial climate states. However, the land ice record is an imperfect tracker, as it is insensitive to changes in temperatures during non-glacial intervals and is influenced by additional factors such as the physical geography of the continents during glacial intervals. To seek to deal with these issues, we take two approaches for comparison. The first to to look at the correlation between LIP area and the extent of ice away from the pole (Fig. \ref{fig:LIP_correlation}). The second is to consider the overall correlation between intervals of high LIP area in the tropics (defined as LIP area in the tropics \textgreater30\% of the maximum) and intervals of glacial climate (defined as ice extent \textgreater30\% of the maximum). This latter approach places less emphasis on the specific magnitudes of the compiled ice extent and LIP area records.

We note that another approach that has been implemented in the literature (e.g. \citealp{Johansson2018a}) is to compare LIP area records to proxy compilations of pCO$_2$. Such proxy compilations are problematic as they are poorly calibrated in deep time, prone to alteration, and reliant on a myriad of assumptions. We thus prefer to use the latitudinal extent of land ice to reflect Earth's overall climate state.

@@

In the original models that proposed the ``Fire and Ice'' hypothesis as an explanation for the onset of Sturtian glaciation, chemical weathering was modeled as an Arhennius relationship where it was solely a function of temperature and run-off \citep{Donnadieu2004a}. However, such an approach neglects the effects that can result through soil shielding and regolith development in low-relief regions. Recent progress on understanding the relationships between landscapes, topography and chemical weathering reveals that these effects are quite important (e.g. \cite{Maher2014a}). As a result, more recent modeling of chemical weathering that has incorporated such processes highlights the importance of high-relief regions relative to low-relief ones \citep{Godderis2017b}. Without active uplift, soil shielding from regolith development on low-relief LIPs could significantly decrease the local weatherability of a LIP and mute its impact on global weatherability. In contrast, processes that lead to continued exhumation of mafic lithologies and the creation of steep topography, particularly in tropical regions, may exert a strong control on global weatherability and long-term climate. This interpretation underlies the hypothesis that arc-continent collisons in the tropics during the Ordovician \citep{Swanson-Hysell2017a} and the Cenozoic \citep{Jagoutz2016a} played a significant role in transitions into glacial climate states at those times.

A complication with the interpretation of soil shielding and limited weathering of LIPs is the rapid area decay rate inferred from the comparison of current LIP extent to estimated original extent. A couple considerations are relevant with respect to this analysis: 1) the current extent of LIP exposure is reduced in part by volcanics being covered by unconsolidated sediments (i.e. regolith development itself) in a number of the provinces; 2) the initial LIP extents are typically poorly constrained and are likely overestimates which could be resulting in inflated interpreted decay rates. Future efforts to develop better-constrained estimates of original LIP extent will improve analyses such as that in this contribution.

We have focused this analysis on the Phanerozoic record given that well-constrained paleogeographic models are available for the past $\sim$500 Myr. Furthermore, the approach of seeking to evaluate correlation between LIP area and glaciation is complicated for Neoproterozoic pan-glacial events wherein ice-albedo runaway leads to persistent global glaciation absent of continued forcing through normal carbon cycle processes until the build-up of sufficient pCO$_2$ to drive deglaciation. Nevertheless, the  hypothesis of tropical LIP area associated with the ca. 720 Ma Franklin LIP increasing global weatherability and contributing to the onset of the Sturtian Snowball Earth glaciation is a major motivating driver behind conducting this analysis. How does the tropical LIP area associated with the Franklin LIP compare to that observed in the Phanerozoic? DISCUSS FRANKLIN AND UMKONDO AREAS AND COMPARE TO THE PHANEROZOIC RECORD.

@@

\begin{figure}[h!]
\begin{center}
	\includegraphics[width=\textwidth]{Manuscript/Figures/LIP_Map.pdf}
	\caption{\textbf{Map of current extent of lithologies associated with LIPs that erupted between 520 Ma and the present (shown in orange) as well as the estimates of the initial LIP area used in the area analysis (shown in blue and slightly modified from \citealp{Ernst2017a}).}}
	\label{fig:LIP_map}
\end{center}
\end{figure}
